---
title: "Machine Learning Course Project"
author: "Zachery Dickson"
date: "January 30, 2016"
output: html_document
---

```{r REQUIREMENTS, message = FALSE}
     library(caret)
     library(foreach)
```

##Introduction
     With the advent of various personal activity monitoring technologies it is now possible to easily attain large amounts of information about body movements whcih can be analyzed and used in a wide variety of applications. This project will use data collected by Ugulino et al. on body movements as measured with accelerometers on a belt, the arm, a handheld dumbell, and the forearm. Six volunteers were measured performing barbell lifts correctly and incorrectly in five different ways. The focus of this report is to use the data to develop a prediction model to classify the movements into each lifting technique given the accelerometer data.

##Exporatory Analysis
```{r LOADING, cache=TRUE}
     trainingUrl = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
     testingUrl = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
     tmpfile <- tempfile()
     download.file(trainingUrl,tmpfile)
     d.train <- read.csv(tmpfile)
     download.file(testingUrl,tmpfile)
     d.test <- read.csv(tmpfile)
     remove(testingUrl,trainingUrl,tmpfile)
```

     The training dataset contains 160 variables: 7 measurement descriptors, 38 measurements for each of the four accelerometer locations, and the class of lifting technique.
     Of the seven descriptors, three are the time the measurement was taken, and X is simply an ID number for each measurement. The user.name varible is the individual performing the lift for that measurement. The num.window and new.window variables describe the measurement period over which the measurements were taken. For this analysis only the user.name variable is likely to have predictive value, as the other variables are arbitrary, specific to this dataset, and would likely decrease the out of sample accuracy.
     Of the 38 measurement variables 25 of have values for only 406 of the entries, this is because these variables are summarries about a measurement window rather than instantaneous measurements. However, there are 858 unique measurement window numbers. the entries with summaries correspond with those entries where the new.window variable is equal to yes. It is likely that either in the original data, or in the construction of the training set, some yes values were excluded. Upon further examination it also appears that many of the summaries are incorrectly calculated. For example, what is listed as the maximum roll on the belt accelerometer for window 12 (-94.3) is acctually the maximum yaw on the belt for that window. This pattern of being off by a few columns is repeated throughout the dataset. 
     The summaries for measurement windows would likely be better for predictions, capturing the entire movement in fewer data points, rather than instantaneous measurements. However, as a result of the deficiencies listed above, and the testing dataset not having any summary data, the summaries will not be included in the model. Insted new summaries will be generated for each window number in the dataset. The median will be used as the sitribution on several of the measurements within the window is skewed.

```{r SUBSET, cache = TRUE}
     cases <- foreach(i = 1:ncol(d.train), .combine = c) %do% {
          sum(!is.na(d.train[,i]) & d.train[,i] != "")
     }
     inBuild <- cases != sum(d.train$new_window == "yes")
     inBuild[c(1,3:6)] = FALSE # Remove non User Descriptor variables
     d.train2 <- d.train[order(d.train$num_window),inBuild]
     d.train3 <- aggregate(d.train2[,-c(1,55)],list(d.train2$num_window), median)
     firstInst <- match(unique(d.train2$num_window),d.train2$num_window)
     d.train3$user_name = d.train2$user_name[firstInst]
     d.train3$classe = d.train2$classe[firstInst]
     d.train3$Group.1 <- NULL
     remove(cases,inBuild, i, firstInst)
```

After removing the summary variables all of the entries are complete. and there are 54 variables: The user.name, the classe, and 13 instantaneous meausurements for each the belt, arm, forearm, and dumbell. The entries are mostly evenly divided between the 6 users, with Pedro having the fewest and Adelmo having the most. Classe A has the most entries by far with the remainder relatively evenly diveded between classes B through E.

```{r SUMMARY}
     summary(d.train3$user_name)
     summary(d.train3$classe)
```

## Model Building

     For this report we will develop a model using the random forest algorithm. The model's out of sample accuracy will be assessed by cross validating with 10-k folds of the training set. The data will first be permuted such that each fold contains each of the outcomes. The validation develops the model based on the median values for each accelerometer measurement window, the accuracy estimate is based on the remaining windows in the complete dataset left out of the model training. 
     
```{r BUILD, message = FALSE, cache = TRUE}
     set.seed(713)
     crossValidate.windows <- function(className,X,Y,method,K) {
          X <- X[sample(1:nrow(X)),]
          foldLength <- nrow(X) / K
          AccuracyEstimate <- numeric()
          for(k in 1:K){
               foldRange <- (foldLength * (k-1) + 1):(foldLength * (k))
               test.windows <- X$num_window[foldRange]
               model.fit <- train(X[-foldRange,className] ~ . ,method = method, data = X[-foldRange,])
               test.Y <- Y$num_window %in% test.windows
               pred <- predict(model.fit,newdata = Y[test.Y,])
               AccuracyEstimate <- c(AccuracyEstimate, sum(pred == Y[test.Y,className]) / length(pred) *100)
          }
          mean(AccuracyEstimate)
     }
     AccEst_RF <- crossValidate.windows("classe",d.train3,d.train2,"rf",10)
     RF.model <- train(classe ~. , method = "rf", data = d.train3)
```

     The estimated out-of-sample accuracy is the mean of the accuracy for each of the 10 folds.

```{r ACCURACY, echo=FALSE}
     AccEst_RF
```

## Prediction

     Applying this model to the test data set results in 19/20 or 95% accuracy, as determined by a single attempt at submitting the predictions.
```{r PREDICTION, message=FALSE}
     predict(RF.model,newdata = d.test)
```


##References

Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6.